A comparative analysis of AquaMaps and IUCN species ranges

### 2015-12-09 data_explore_2015_data.Rmd
added R Markdown codes in manuscript to calc numbers etc

dir_anx <- file.path(dir_N, 'git-annex/globalprep/SPP_ICO')
dir_data <- file.path(dir_N, 'git-annex/globalprep/_raw_data')
dir_fig <- file.path(getwd(), 'figs2015')

>>>>

dir_anx <- file.path(dir_N, 'git-annex/globalprep/SPP_ICO')
dir_data <- file.path(dir_N, 'git-annex/globalprep/_raw_data')
dir_fig <- file.path(getwd(), 'figures')

### 2015-12-09 ms_outline.Rmd
added R Markdown codes in manuscript to calc numbers etc

##Main message

@@

##Working Abstract  
<!--
* Importance of understanding spatial distribution of species
    * help identify hotspots for global biodiversity 
    * priorities for conservation
    * marine spatial planning
    * assessing biodiversity 
    * monitor impacts (human, natural)
    * global-scale assessments
    * poor overlap of species lists between the two data sets - sometimes helpful to use both
* Challenges
    * difficulty of monitoring marine species in space
    * even more difficult to monitor over time
        * can Aquamaps, as a bioclimatic envelope model, be used to predict changes or used as proxy for species shifts if envelopes are known to have shifted? __can we use this as trend? or does all that count as pressures? seems like human factors can drastically change the trend that wouldn't be captured here__
        * Does that additional data in Aquamaps include predictions for 2050, 2100 (bioclimatic envelopes) - how is that reflected in the species distributions?
* IUCN vs Aquamaps
    * spatial distribution efforts range from modeled to expert input
    * distribution data for 1900+ species found in both data sets
    * differences in predictions of species presence in both distribution and area
        * Aquamaps validation and benefits from _Ready et al. 2010_:
          *  We present and evaluate AquaMaps, a presence-only species distribution modelling system that allows the incorporation of expert knowledge about habitat usage and was designed for maximum output of standardized species range maps at the global scale. In the marine environment there is a significant challenge to the production of range maps due to large biases in the amount and location of occurrence data for most species. AquaMaps is compared with traditional presence-only species distribution modelling methods to determine the quality of outputs under equivalently automated conditions.  _The effect of the inclusion of expert knowledge to AquaMaps is also investigated._ 
          * Model outputs were tested internally, through data partitioning, and externally against independent survey data to determine the ability of models to predict presence versus absence. Models were also tested externally by assessing correlation with independent survey estimates of relative species abundance. AquaMaps outputs compare well to the existing methods tested, and inclusion of expert knowledge results in a general improvement in model outputs. The transparency, speed and adaptability of the AquaMaps system, as well as the existing online framework which allows expert review to compensate for sampling biases and thus improve model predictions are proposed as additional benefits for public and research use alike.
-->

@@

#Introduction

@@

The IUCN Red List of Threatened Species includes more than 19,000 assessed marine species, 
<!--define what "assessed" means? include DD? keeping those in for now--> but only **4463** <!--11975 maps in those shapefiles, but fewer than 4500 correspond to assessed marine species... and only 2500 of those overall correpond to AquaMaps species--> of these have associated spatial information. AquaMaps has developed range maps for over 22,800 species. Only 2,100 species have been mapped in both datasets. This gap in overlapping species between these two datasets further highlights the need for understanding each of these datasets. Both IUCN and AquaMaps have gaps in marine species information and a significant portion of available data may be overlooked if limited to just one of these datasets.

@@

IUCN only publishes spatial distribution map sets for taxonomic groups in which at least 90% of the species within the taxon group have been evaluated.  While this helps prevent sampling bias within taxa, it also means that entire taxonomic groups remain unavailable until they have met this threshold of comprehensive assessment.  As of December 2015, IUCN species distribution maps were available for XXX marine species (habitat = "marine" on IUCN species information pages).

@@

The IUCN publishes species distribution maps for as collections of shapefiles for a range of taxonomic groups.  Species spatial distributions are determined by experts based upon <known occurrence, suitable habitat, and blah>.  The IUCN presents these maps as 'limits of distibution' with the caveat that the species 'probably occurs within this polygon, but it does not mean that it is distributed equally within that polygon or occurs everywhere within that polygon'.<cite?>  IUCN only publishes spatial distribution map sets for taxonomic groups in which at least 90% of the species within the taxon group have been evaluated.  While this helps prevent sampling bias within taxa, it also means that entire taxonomic groups remain unavailable until they have met this threshold of comprehensive assessment.  As of December 2015, IUCN species distribution maps were available for XXX marine species (habitat = "marine" on IUCN species information pages).

@@

AquaMaps (Kaschner et al. 2013) offers species range maps based on modeled relative environmental suitability.  For each assessed species, data from observations and expert knowledge are used to generate bioclimatic envelopes for sea surface temperature, salinity, depth, productivity, and where applicable, sea ice cover and/or distance to land.  The species' suite of bioclimatic envelopes is compared to a map of environmental attributes on a 0.5 degree global grid, creating a global map of ___probability of occurrence___ (or ___relative environmental suitability?___ (see Kaschner et al 2006 and 2011)), with cell values ranging from 0 (unsuitable) to 1.00 (highly suitable).  These maps can then be tweaked by experts altering the parameters and regenerating the modeled map. As of December 2015, ___22,000___ species have been mapped. ___Expert mode to further refine models?___ See (Kaschner et al. 2013) for more detailed methods.

@@

___number of species in each data set, and the number that overlap___

@@

Researchers have two large and useful data sets at their disposal, each with its own utility.  Being able to use both simultaneously makes many more species available, but here there be challenges.

@@

* Definition of "alignment" in two dimensions, incl "good" vs "poor"
* basic methodology (raster, overlay, calculate)? or should there be a separate methodology section?

<!--
* Here we can discuss the idea of multiple "dimensions" of alignment - distributional and area alignment.  Well-aligned species will show high alignment in both dimensions.  
* Maybe a diagram here?
* What can alignment tell us about the individual species, the species groups, and the data sets?
* Note that alignment does not necessarily mean accuracy or precision - a poorly-understood species could appear to be well-aligned.  
-->

@@

To facilitate our analysis, we selected the subset of marine species that have range maps in both IUCN and AquaMaps (n = ???), matched by Genus and species names, and then rasterized the IUCN shapefiles to match the Aquamaps half-cell raster.  We flattened each map by assigning to each cell a simple "presence/absence" value:

* For IUCN data, any cell with non-zero overlap with a range map polygon was considered "presence."  
* For AquaMaps data, any cell with non-zero probability (or RES) was considered "presence."  
* In both cases, these criteria for presence will maximize the number of cells included.  Using these criteria, we found that for 72% ??? of species, the IUCN distribution maps indicated a larger species range than the range indicated by AquaMaps.

@@

For each of the ___2200___ species whose range is described by both data sets, we calculated the distribution match and the area ratio, and plotted the ___2200___ species map pairs on a scatter plot of distribution alignment vs area alignment (Figure 1).

@@

___Figure 1: Distribution alignment vs area alignment for ___2200___ species mapped in both Aquamaps and IUCN species distribution maps.___

@@

| Description              |     Distribution match    |    Area match     |
| :----------------------: | :-----------------------: | :---------------: |
| q1: Well-aligned         | Distribution match > XX % | Area match > YY % |
| q2: Distribution-aligned | Distribution match < XX % | Area match > YY % |
| q3: Area-aligned         | Distribution match > XX % | Area match < YY % |
| q4: Poorly-aligned       | Distribution match < XX % | Area match < YY % |

Clearly the upper right quadrant (quadrant 1) is the place to be.  Species in this quadrant generally agree in spatial distribution between the two distribution maps, and the area of described ranges will be similar; expert-drawn range map and bioclimatic envelope model produce similar predictions of species distribution. (n = XXX; YYY %)

The other quadrants each suggest a different story.  Species map pairs that fall within the upper left quadrant (quadrant 2: well-matched in distribution but poorly-matched in area) indicate species in which the range of the smaller map falls generally within the larger map, but the larger map may include more generous buffers or may include areas unrepresented in the smaller map. (n = XXX; YYY %)

The lower right quadrant (quadrant 3: well-matched in area but poorly matched in distribution) includes species for which the two maps generally agree in range area, but disagree in where those ranges occur.  General disagreement in distribution seems to indicate a greater issue than general disagreement in area extent, so we categorized map pairs in this quadrant as being more problematic than map pairs in either of the upper quadrants. (n = XXX; YYY %)

The lower left quadrant (quadrant 4: poorly matched in both dimensions) indicates species for which the map pairs fail to agree in both area and distribution. (n = XXX; YYY %)

@@

Figure 2 breaks the ___2200___ species map pairs into taxonomic groups as classified by the published IUCN range maps.  We were not surprised to find that map pairs for marine mammals, tunas, and billfishes were well-matched (quadrants 1 and 2); many of these species are global-ranging pelagic organisms, with some ranges encompassing nearly the entire ocean.  Mangroves are also not a surprise, being immobile and easily mapped by remote sensing methods.

@@

![](figures/scatter_AMThreshold.png)
Effect of selecting different thresholds on total "present" area - linearish, no big steps or drops to indicate sudden change in tradeoff.

@@

Halpin, P.N., A.J. Read, E. Fujioka, B.D. Best, B. Donnelly, L.J. Hazen, C. Kot, K. Urian, E. LaBrecque, A. Dimatteo, J. Cleary, C. Good, L.B. Crowder, and K.D. Hyrenbach. 2009. OBIS-SEAMAP: The world data center for marine mammal, sea bird, and sea turtle distributions. Oceanography 22(2):104-115

>>>>

``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}

# library(ggplot2)
# library(maps)
# library(readr)
# library(raster)
# library(data.table)
library(tidyr)
library(dplyr)
library(stringr)

# dir_N <- c('Windows' = '//neptune.nceas.ucsb.edu/data_edit',
#            'Darwin'  = '/Volumes/data_edit',
#            'Linux'   = '/var/data/ohi')[[ Sys.info()[['sysname']] ]]

# dir_anx <- file.path(dir_N, 'git-annex/globalprep/SPP_ICO')
# dir_data <- file.path(dir_N, 'git-annex/globalprep/_raw_data')
# dir_fig <- file.path(getwd(), 'figures')
```

``` {r load data, echo = FALSE, warning = FALSE, message = FALSE}
### use data_explore.Rmd to create this file, if it does not yet exist.

data_file <- 'data/spp_list_w_area_percent_noextinct_2015.csv'
spp_map_pairs <- read.csv(data_file, stringsAsFactors = FALSE) %>%
  filter(!is.na(popn_category)) %>% ### filter out unassessed species
  filter(!is.na(sm_perc))           ### filter out NA maps (usually IUCN maps with no polygons)

### generate other lists

spp_all_file <- 'data/spp_all_2015.csv'
spp_all <- read.csv(spp_all_file, stringsAsFactors = FALSE)

iucn_maps_file <- 'data/spp_iucn_maps.csv'
spp_iucn_maps <- read.csv(iucn_maps_file, stringsAsFactors = FALSE)

iucn_marine_file <- 'data/spp_iucn_marine.csv'
spp_iucn_marine <- read.csv(iucn_marine_file, stringsAsFactors = FALSE)

spp_am_file <- 'data/spp_am.csv'
spp_am <- read.csv(spp_am_file, stringsAsFactors = FALSE)

```
## Main message

@@

<!--
##Working Abstract  
* Importance of understanding spatial distribution of species
    * help identify hotspots for global biodiversity 
    * priorities for conservation
    * marine spatial planning
    * assessing biodiversity 
    * monitor impacts (human, natural)
    * global-scale assessments
    * poor overlap of species lists between the two data sets - sometimes helpful to use both
* Challenges
    * difficulty of monitoring marine species in space
    * even more difficult to monitor over time
        * can Aquamaps, as a bioclimatic envelope model, be used to predict changes or used as proxy for species shifts if envelopes are known to have shifted? __can we use this as trend? or does all that count as pressures? seems like human factors can drastically change the trend that wouldn't be captured here__
        * Does that additional data in Aquamaps include predictions for 2050, 2100 (bioclimatic envelopes) - how is that reflected in the species distributions?
* IUCN vs Aquamaps
    * spatial distribution efforts range from modeled to expert input
    * distribution data for 1900+ species found in both data sets
    * differences in predictions of species presence in both distribution and area
        * Aquamaps validation and benefits from _Ready et al. 2010_:
          *  We present and evaluate AquaMaps, a presence-only species distribution modelling system that allows the incorporation of expert knowledge about habitat usage and was designed for maximum output of standardized species range maps at the global scale. In the marine environment there is a significant challenge to the production of range maps due to large biases in the amount and location of occurrence data for most species. AquaMaps is compared with traditional presence-only species distribution modelling methods to determine the quality of outputs under equivalently automated conditions.  _The effect of the inclusion of expert knowledge to AquaMaps is also investigated._ 
          * Model outputs were tested internally, through data partitioning, and externally against independent survey data to determine the ability of models to predict presence versus absence. Models were also tested externally by assessing correlation with independent survey estimates of relative species abundance. AquaMaps outputs compare well to the existing methods tested, and inclusion of expert knowledge results in a general improvement in model outputs. The transparency, speed and adaptability of the AquaMaps system, as well as the existing online framework which allows expert review to compensate for sampling biases and thus improve model predictions are proposed as additional benefits for public and research use alike.
-->

@@

# Introduction

@@

The IUCN Red List of Threatened Species includes more than 19,000 assessed marine species, 
<!--define what "assessed" means? include DD? keeping those in for now--> but fewer than 4,500 <!--11975 maps in those shapefiles, but fewer than 4500 correspond to assessed marine species... and only 2500 of those overall correpond to AquaMaps species--> of these have associated spatial information. AquaMaps has developed range maps for over 22,800 species. Only 2,200 species are mapped in both datasets. This gap in overlapping species between these two datasets further highlights the need for understanding each of these datasets. Both IUCN and AquaMaps have gaps in marine species information and a significant portion of available data may be overlooked if limited to just one of these datasets.

@@

IUCN only publishes spatial distribution map sets for taxonomic groups in which at least 90% of the species within the taxon group have been evaluated.  While this helps prevent sampling bias within taxa, it also means that entire taxonomic groups remain unavailable until they have met this threshold of comprehensive assessment.  As of December 2015, IUCN species distribution maps were available for XXX marine species in YYY taxonomic groups.  However, only YYY of these species have been formally assessed under the IUCN Red List of Threatened Species (listed on that spreadsheet, and habitat = "marine" on IUCN species information pages).

@@

The IUCN publishes species distribution maps for as collections of shapefiles for a range of taxonomic groups.  Species spatial distributions are determined by experts based upon ___known occurrence, suitable habitat, and blah___.  The IUCN presents these maps as 'limits of distibution' with the caveat that the species 'probably occurs within this polygon, but it does not mean that it is distributed equally within that polygon or occurs everywhere within that polygon'.<cite?>  IUCN only publishes spatial distribution map sets for taxonomic groups in which at least 90% of the species within the taxon group have been evaluated.  While this helps prevent sampling bias within taxa, it also means that entire taxonomic groups remain unavailable until they have met this threshold of comprehensive assessment.  As of December 2015, IUCN had published species distribution maps for nearly 12,000 marine species across 26 taxonomic groups, though fewer than 4,500 of these have been formally assessed under the IUCN Red List of Threatened Species.

@@

AquaMaps (Kaschner et al. 2013) offers species range maps based on modeled relative environmental suitability.  For each assessed species, data from observations and expert knowledge are used to generate bioclimatic envelopes for sea surface temperature, salinity, depth, productivity, and where applicable, sea ice cover and/or distance to land.  The species' suite of bioclimatic envelopes is compared to a map of environmental attributes on a 0.5 degree global grid, creating a global map of ___probability of occurrence___ (or ___relative environmental suitability?___ (see Kaschner et al 2006 and 2011)), with cell values ranging from 0 (unsuitable) to 1.00 (highly suitable).  These maps can then be tweaked by experts altering the parameters and regenerating the modeled map. As of December 2015, nearly 23,000 species have been mapped, though only 5,600 of these have been evaluated for the IUCN Red List. ___Expert mode to further refine models?___ See (Kaschner et al. 2013) for more detailed methods.

@@

| Map set   | n~total~                | n~Red_List~                                                               |
| :-------: | ----------------------: | ------------------------------------------------------------------------: |
| AquaMaps  | `r nrow(spp_am)`        | `r nrow(spp_am %>% filter(sciname %in% spp_iucn_marine$sciname))`         |
| IUCN      | `r nrow(spp_iucn_maps)` | `r nrow(spp_iucn_maps %>% filter(binomial %in% spp_iucn_marine$sciname))` |
| Both      | `r nrow(spp_am %>% filter(sciname %in% spp_iucn_maps$binomial))` | `r nrow(spp_map_pairs)`          |

@@

Researchers have two large and useful data sets at their disposal, each with its own utility.  Being able to use both simultaneously makes many more species available, but beware: here there be challenges.

@@

<!--
___Main messages:___

* Definition of "alignment" in two dimensions, incl "good" vs "poor"
* basic methodology (raster, overlay, calculate)? or should there be a separate methodology section?
-->

<!--
* Note that alignment does not necessarily mean accuracy or precision - a poorly-understood species could appear to be well-aligned.  
-->

@@

To facilitate our analysis, we selected the subset of marine species that have range maps in both IUCN (filtering out ranges identified as extinct) and AquaMaps current native distribution (n = `r nrow(spp_map_pairs)`), and used genus and species binomials to identify paired maps.  We then rasterized the IUCN shapefiles to match the Aquamaps half-cell raster.  We flattened each map by assigning to each cell a simple "presence/absence" value:

* For IUCN data, we defined "presence" within a cell as any non-zero overlap with a range map polygon.
* For AquaMaps data, we defined presence as any cell with non-zero probability ___(or RES?)___.  
* In both cases, these criteria for presence will maximize the number of cells included.  Using these criteria, we found that for 
`r round(nrow(spp_map_pairs %>% filter(sm_range == 'AM')) / nrow(spp_map_pairs) * 100, 1)`%
of species, the IUCN distribution maps indicated a larger species range than the range indicated by AquaMaps.

@@

For each of the `r nrow(spp_map_pairs)` species whose range is described by both data sets, we calculated the distribution match and the area ratio, and plotted the species map pairs on a scatter plot of distribution alignment vs area alignment (Figure 1).

@@

___Figure 1: Distribution alignment vs area alignment for `r nrow(spp_map_pairs)` species mapped in both Aquamaps and IUCN species distribution maps.___

@@

``` {r calc medians and quadrant counts, echo = FALSE}
dist_median <- round(median(spp_map_pairs$sm_perc,    na.rm = TRUE), 1)
area_median <- round(median(spp_map_pairs$area_ratio, na.rm = TRUE), 1)
q1_count <- sum(spp_map_pairs$sm_perc > dist_median  & spp_map_pairs$area_ratio >  area_median)
q2_count <- sum(spp_map_pairs$sm_perc <= dist_median & spp_map_pairs$area_ratio >  area_median)
q3_count <- sum(spp_map_pairs$sm_perc > dist_median  & spp_map_pairs$area_ratio <= area_median)
q4_count <- sum(spp_map_pairs$sm_perc <= dist_median & spp_map_pairs$area_ratio <= area_median)
```

| Description              |   Distribution match |      Area match      |
| :----------------------: | :------------------: | :------------------: |
| q1: Well-aligned         |  > `r dist_median` % |  > `r area_median` % |
| q2: Distribution-aligned |  < `r dist_median` % |  > `r area_median` % |
| q3: Area-aligned         |  > `r dist_median` % |  < `r area_median` % |
| q4: Poorly-aligned       |  < `r dist_median` % |  < `r area_median` % |

The upper right quadrant (quadrant 1) comprises species whose maps largely agree (better than median value) in both spatial distribution and the area of described ranges.  For these species, the expert-drawn range map and the bioclimatic envelope model produce similar predictions of species distribution. 
(n = `r q1_count`; `r round(q1_count/nrow(spp_map_pairs) * 100, 1)` %)

The other quadrants each suggest a different story.  Species map pairs that fall within the upper left quadrant (quadrant 2: well-matched in distribution but poorly-matched in area) indicate species in which the range of the smaller map falls generally within the larger map, but the larger map may include more generous buffers or may include areas unrepresented in the smaller map. 
(n = `r q2_count`; `r round(q2_count/nrow(spp_map_pairs) * 100, 1)` %)

The lower right quadrant (quadrant 3: well-matched in area but poorly matched in distribution) includes species for which the two maps generally agree in range area, but disagree in where those ranges occur.  Disagreement in distribution seems to indicate a greater issue than disagreement in area extent, so we categorized map pairs in this quadrant as being more problematic than map pairs in either of the upper quadrants. 
(n = `r q3_count`; `r round(q3_count/nrow(spp_map_pairs) * 100, 1)` %)

The lower left quadrant (quadrant 4: poorly matched in both dimensions) indicates species for which the map pairs fail to agree in both area and distribution. 
(n = `r q4_count`; `r round(q4_count/nrow(spp_map_pairs) * 100, 1)` %)

@@

Figure 2 divides the `r nrow(spp_map_pairs)` species map pairs into taxonomic groups as classified by the published IUCN range maps, and examines the breakdown of each taxonomic group by quadrant.

___move to discussion___ We were not surprised to find that map pairs for marine mammals, tunas, and billfishes were well-matched (quadrants 1 and 2); many of these species are global-ranging pelagic organisms, with some ranges encompassing nearly the entire ocean.  Mangroves are also not a surprise, being immobile and easily mapped by remote sensing methods.

@@

![](figures/scatter_AMThreshold.png)
Effect of selecting different thresholds on total "present" area - linearish, no big steps or drops to indicate sudden change in tradeoff.  Do regression - slope indicates that sensitivity is less than 1:1

@@

<!--
Halpin, P.N., A.J. Read, E. Fujioka, B.D. Best, B. Donnelly, L.J. Hazen, C. Kot, K. Urian, E. LaBrecque, A. Dimatteo, J. Cleary, C. Good, L.B. Crowder, and K.D. Hyrenbach. 2009. OBIS-SEAMAP: The world data center for marine mammal, sea bird, and sea turtle distributions. Oceanography 22(2):104-115
-->
